---
title: "COVID-19 Pandemic Analysis: Does Income Affect Mortality Rate?"
author: "Vatanak Lavan"
date: "09/05/2022"
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
---

# Initialize

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Local Environment

The package `renv` is installed to the project to maintain a reproducible workflow in RStudio. `renv` allows for the creation of a local environment with a private library to manage the project's R dependencies.

```{r}
# install.packages("renv")

# Initialize new local environment with private library
# renv::init()
```

## Tidyverse

The `tidyverse` is an opinionated collection of R packages designed for data science. All of its packages share an underlying design philosophy, grammar, and data structures.

Each one has tools created specifically for different points throughout the process. Utilizing `tidyverse`, specific steps are followed in undertaking a data science project.

```{r message=FALSE, warning=FALSE}
# install.packages("tidyverse")  
library(tidyverse)
```

Many functions used throughout the programming of the project to manipulate data are provided by the `dplyr` package. It is called often to extract, modify and combine tibbles based on specified conditions.

# Wrangle

## Import

### Connections

The package `odbc` is installed to setup connections to the databases in RStudio via Microsoft ODBC (Open Database Connectivity) drivers. It serves as a backend which the `DBI` package uses to define a frontend communication interface in RStudio.

```{r}
# install.packages("odbc")
library(odbc)

# List available ODBC User data sources on system
# odbcListDataSources()
```

`odbc` requires the installation of MySQL Connector/ODBC. ODBC User data sources are added through Microsoft ODBC Data Source Administrator using MySQL ODBC 8.0 ANSI Driver. Connection parameters are specified in MySQL Connector/ODBC.

Connections are created from User DSNs (Data Source Name) by specifying its name.

```{r}
# Create connection to simplemaps counties database 
con_us_counties <- dbConnect(odbc(), "simplemaps")

# Create connection to 2021spfdb COVID-19 deaths database
con_covid19_deaths <- dbConnect(odbc(), "2021spfdb")
```

### Queried Tables

Tables fetched from the submitted SQL queries are imported as tibbles. Tibbles are a table format provided by the `tibble` package. They inherit R's traditional data.frame class, but have improved behaviors.

```{r}
# Create SQL query string 
query_us_counties <- "SELECT * from uscounties"

# Fetch queried table from as tidyr tibble
us_counties <- dbGetQuery(con_us_counties,query_us_counties) %>% 
  as_tibble()

us_counties
```

The `map_id` column in the county_data table contain FIPS (Federal Information Processing Standards) codes of each corresponding county. It is renamed as county_fips to match that of the uscounties table.

Rows are retrieved where the date value is `2022-05-13` to maintain accuracy. This is because the data in the simplemaps United States Counties Database was updated as of May 13, 2022.

```{r}
# Create SQL query string
query_covid19_deaths <- "SELECT map_id as county_fips, cases, deaths FROM county_data WHERE date='2022-05-13'"

# Fetch queried table as tidyr tibble
covid19_deaths <- dbGetQuery(con_covid19_deaths, query_covid19_deaths) %>%
  as_tibble()

covid19_deaths
```

```{r}
dbDisconnect(con_us_counties)
dbDisconnect(con_covid19_deaths)

rm(con_us_counties, con_covid19_deaths, query_us_counties, query_covid19_deaths)
```

### Excel Spreadsheet

We also utilize the the national-level median household income from the U.S. Census Bureau as a means of categorizing a county's own median household income in comparison to that of the whole country.

```{r}
library(readxl)

us_states_income <- read_xlsx(
  "us_census_median_household_income.xlsx",
  col_names = c("state_name","median_income_state"),
  range = "A10:B60"
  )

us_states_income
```

## Tidy

### Excessive Fields

The simplemaps United States Counties Database contains seventy-eight fields for each county. However, a number of these are considered irrelevant and are not included.

```{r}
us_counties_income <- us_counties %>%
  select(
    county_fips,
    county,
    state_id,
    state_name,
    population,
    density,
    income_household_median
  )

us_counties_income
```

### US Territories

The simplemaps United States Counties Database includes data from the US territories - American Samoa, Guam, Northern Mariana Islands, Puerto Rico, and Virgin Islands.

This is excluded because the `county_data` table contains data excluding that of US territories. Additionally, the data for US territories in the `uscounties` table is filled with missing values.

```{r}
# Create list of state_id values of US territories
territory_ids <- c("AS","GU","MP","PR","VI")

# Create tibble of US territories
us_counties_income %>% 
  filter(state_id %in% territory_ids) 
```

```{r}
# Create %notin% function 
`%notin%` <- Negate(`%in%`)

# Create tibble of US states
us_counties_income <- us_counties_income %>%
  filter(state_id %notin% territory_ids)

us_counties_income
```

```{r}
rm(territory_ids)
```

### Missing Values

Handling missing values is mandatory in any data science project. The `tidyr` package offers different approaches to this. However, through inspecting the tibbles, neither of them contain any NA values.

```{r}
# Display tibble of missing values in counties_states
us_counties_income %>% 
  filter(if_any(everything(), is.na))
```

```{r}
# Display tibble of missing values in covid_deaths
covid19_deaths %>% 
  filter(if_any(everything(), is.na))
```

## Transform

### Join

After tidying the tibbles, both will be combined on the `county_fips` column. This is executed through `dplyr`'s `inner_join` function which retains only rows with matches between both.

```{r}
# Merge us_counties and covid19_deaths
us_counties_income_deaths <-
  inner_join(us_counties_income, covid19_deaths, by = "county_fips") %>%
  arrange(county_fips) # Arrange by county_fips column

us_counties_income_deaths
```

```{r}
rm(us_counties, covid19_deaths)
```

### Income

```{r}
# Calculate the death rate per 100,000 population
us_counties_income_deaths <- us_counties_income_deaths %>%
  mutate(death_rate = round(((deaths / population) * 100000), digits = 2)) %>%
  
  # Move the 'death_rate' column next to 'deaths'
  relocate(death_rate, .after = deaths)

us_counties_income_deaths
```

# Analyze

## Classify Categories

```{r}
# Retrieve the national median household income in the 'us_states_income' dataframe
us_income_household_median <- us_states_income$median_income_state[1]

# Calculate the total population across all counties in the 'us_counties_income_deaths' dataframe
us_population <- us_counties_income_deaths %>%
  pull(population) %>% sum()

# Calculate the total number of COVID-19 deaths across all counties in the 'us_counties_income_deaths' dataframe
us_covid19_deaths <- us_counties_income_deaths %>%
  pull(deaths) %>% sum()

# Calculate the COVID-19 death rate per 100,000 population for the entire dataset
us_covid19_death_rate <- (us_covid19_deaths / us_population) * 100000
```

```{r}
rm(us_states_income)
```

```{r}
# Add a new column 'category' with initial NA values to the 'us_counties_income_deaths' dataframe
us_counties_income_deaths <- us_counties_income_deaths %>%
  add_column(category = NA)

# Loop through each row in the 'us_counties_income_deaths' dataframe
for (i in 1:nrow(us_counties_income_deaths)) {
  # Check if the median household income is less than the overall median income
  if (us_counties_income_deaths$income_household_median[i] < us_income_household_median) {
    # Check if the death rate is less than the overall COVID-19 death rate
    if (us_counties_income_deaths$death_rate[i] < us_covid19_death_rate) {
      # Assign the category "Low Income, Low Death Rate" if both conditions are met
      us_counties_income_deaths$category[i] <- "Low Income, Low Death Rate"
    }
    else {
      # Assign the category "Low Income, High Death Rate" if the income condition is met but not the death rate condition
      us_counties_income_deaths$category[i] <- "Low Income, High Death Rate"
    }
  }
  else {
    # Check if the death rate is less than the overall COVID-19 death rate
    if (us_counties_income_deaths$death_rate[i] < us_covid19_death_rate) {
      # Assign the category "High Income, Low Death Rate" if the income condition is not met but the death rate condition is met
      us_counties_income_deaths$category[i] <- "High Income, Low Death Rate"
    }
    else {
      # Assign the category "High Income, High Death Rate" if neither condition is met
      us_counties_income_deaths$category[i] <- "High Income, High Death Rate"
    }
  }
}
```

```{r}
rm(i, us_income_household_median, us_covid19_death_rate)
```

## Plot Data

```{r message=FALSE, warning=FALSE}
# install.packages("plotly")
library(plotly)
```

### Mortality Rate

```{r}
# Create a scatter plot using Plotly
us_counties_scatter <- plot_ly(
  us_counties_income_deaths,
  x = ~income_household_median / 1000,  
  y = ~death_rate, 
  type = 'scatter',
  mode = 'markers',
  text = paste(us_counties_income_deaths$county, us_counties_income_deaths$state_id, sep = ", ")
)

# Customize the layout of the scatter plot
us_counties_scatter <- us_counties_scatter %>%
  layout(
    # title = "Median Household Income vs COVID-19 Death Rate in U.S. Counties (05-13-2022)",
    xaxis = list(title = 'Median Household Income ($K)'),  # X-axis label
    yaxis = list(title = 'COVID-19 Mortality Rate (per 100K)')  # Y-axis label
  )

us_counties_scatter
```

```{r}
# Perform a correlation test between 'income_household_median' and 'death_rate'
cor.test(
  us_counties_income_deaths$income_household_median,
  us_counties_income_deaths$death_rate
)
```

#### Inspect Outliers

```{r}
# Create a box plot for 'income_household_median'
plot_ly(
  y = us_counties_income_deaths$income_household_median,  # Data for the Y-axis
  text = paste(us_counties_income_deaths$county, us_counties_income_deaths$state_id, sep = ", "),
  type = "box",  # Box plot type
  name = "Median Household Income",  # Trace name for the legend
)

# Create a box plot for 'death_rate'
plot_ly(
  y = us_counties_income_deaths$death_rate,  # Data for the Y-axis
  text = paste(us_counties_income_deaths$county, us_counties_income_deaths$state_id, sep = ", "),
  type = "box",  # Box plot type
  name = "COVID-19 Death Rate",  # Trace name for the legend
)
```

```{r}
# Compute the IQR of the household median income
q1_income <- quantile(us_counties_income_deaths$income_household_median, 0.25)  # 1st Quartile
q3_income <- quantile(us_counties_income_deaths$income_household_median, 0.75)  # 3rd Quartile
iqr_income <- q3_income - q1_income  # Interquartile Range (IQR)

# Compute the IQR of the death rate
q1_death <- quantile(us_counties_income_deaths$death_rate, 0.25)  # 1st Quartile
q3_death <- quantile(us_counties_income_deaths$death_rate, 0.75)  # 3rd Quartile
iqr_death <- q3_death - q1_death  # Interquartile Range (IQR)

# Identify potential outliers based on the IQR method
us_counties_outliers_death <- us_counties_income_deaths %>%
  filter(
    death_rate < q1_death - 1.5 * iqr_death |  # Lower bound for death rate outliers
    death_rate > q3_death + 1.5 * iqr_death |  # Upper bound for death rate outliers
    income_household_median < q1_income - 1.5 * iqr_income |  # Lower bound for income outliers
    income_household_median > q3_income + 1.5 * iqr_income  # Upper bound for income outliers
  )

us_counties_outliers_death
```

```{r}
rm(q1_death, q3_death, iqr_death)
```

```{r}
# Filter the 'us_counties_death_income' dataframe to select specific counties which are significant outliers
us_counties_outliers_significant <- us_counties_income_deaths %>%
  filter(county_fips %in% c(36061, 48301, 48311))

us_counties_outliers_significant
```

```{r}
# Create a scatter plot for all counties in blue
us_counties_scatter <- plot_ly(
  data = us_counties_income_deaths,
  x = ~income_household_median / 1000,
  y = ~death_rate,
  type = 'scatter',
  mode = 'markers',
  name = "County",  # Legend label for counties
  showlegend = FALSE  # Hide the legend for this trace
)

# Add red markers for significant outliers
us_counties_scatter <- us_counties_scatter %>%
  add_markers(
    data = us_counties_outliers_significant,
    marker = list(color = "red"),  # Mark outliers in red
    name = "Outlier"  # Legend label for outliers
  )

# Customize the layout of the scatter plot
us_counties_scatter <- us_counties_scatter %>%
  layout(
    # title = "Median Household Income vs COVID-19 Death Rate in U.S. Counties (05-13-2022)",
    xaxis = list(title = 'Median Household Income ($K)'),  # X-axis label
    yaxis = list(title = 'COVID-19 Mortality Rate (per 100K)')  # Y-axis label
  )

us_counties_scatter
```

# Visualize

## Choropleth

```{r}
# install.packages("rjson")
library(rjson)
```

```{r}
# Load GeoJSON data from a URL
us_counties_geojson <- fromJSON(file = "https://raw.githubusercontent.com/plotly/datasets/master/geojson-counties-fips.json")

# Configure settings for the choropleth map
us_counties_choropleth_config <- list(
  scope = 'usa',  # Set the map scope to USA
  projection = list(type = 'albers usa'),  # Set the map projection type
  showlakes = TRUE,  # Show lakes on the map
  lakecolor = toRGB('white')  # Set the color for lakes
)
```

```{r}
library(RColorBrewer)
```

```{r}
# Convert the 'category' column to a factor
us_counties_income_deaths <- us_counties_income_deaths %>%
  mutate(category = factor(category)) %>% 

# Create a new column 'category_numeric' and store the numeric representation of 'category'
  mutate(category_numeric = as.numeric(category))

us_counties_income_deaths
```

```{r}
# Generate a color palette using Brewer palette 'Set1'
foo <- brewer.pal(n = 6, name = "Set1")[c(6,3,1,2)]

# Assign names to the colors based on the levels of the 'category' column in 'us_counties_income_deaths'
names(foo) = levels(us_counties_income_deaths$category)

# Define a function to calculate breaks for the color scale
Z_Breaks = function(n) {
  CUTS = seq(0, 1, length.out = n + 1)
  rep(CUTS, ifelse(CUTS %in% 0:1, 1, 2))
}

# Create a data frame for the color scale
colorScale <- data.frame(
  z = Z_Breaks(4),  # Breaks for the color scale
  col = rep(foo, each = 2),  # Repeating the colors
  stringsAsFactors = FALSE  # Ensure columns are not treated as factors
)
```

```{r}
# Create a choropleth map
us_counties_choropleth <- plot_ly(
  data = us_counties_income_deaths,  # Data source
  type = "choropleth",  # Choropleth map type
  geojson = us_counties_geojson,  # GeoJSON data for US counties
  locations = ~county_fips,  # Locations to map
  z = ~category_numeric,  # Color scale values
  colorscale = colorScale,  # Custom color scale
  colorbar = list(
    title = list(text = "Category", font = list(size = 16)),  # Colorbar title
    tickvals = c(1.37, 2.13, 2.88, 3.63),  # Tick values
    ticktext = names(foo),  # Tick text labels
    orientation = "h",  # Horizontal colorbar
    y = -0.2,  # Y-position of the colorbar
    reversescale = TRUE  # Reverse the color scale
  ),
  marker = list(line = list(width = 0))  # Set marker line width
)

# Set a title for the choropleth map
# us_counties_choropleth <- us_counties_choropleth %>%
#   layout(title = "U.S. Counties by Median Household Income and COVID-19 Death Rates (05-13-2022)")

# Apply the geo configuration to the choropleth map
us_counties_choropleth <- us_counties_choropleth %>%
  layout(geo = us_counties_choropleth_config)


us_counties_choropleth
```

```{r}
rm(us_counties_geojson, us_counties_choropleth_config)
rm(foo, colorScale)
```
